##(A)

#Load the `penguins` dataset. Select only the columns containing the species, sex, body mass, flipper length and bill length. Remove the rows with missing values in any of these columns.

```{r}
library(palmerpenguins)

df = penguins[,c("species","sex","bill_length_mm","flipper_length_mm","body_mass_g")]
```

```{r}
# omit all rows with any missing values
df = na.omit(df)
```

#visualising the relationship between body mass and the other four variables.

```{r fig.width=5, fig.height=3}

plot(df$flipper_length_mm, df$body_mass_g, 
     xlab = "Flipper Length (mm)",
     ylab="Body Mass (g)")

plot(df$bill_length_mm, df$body_mass_g, 
     xlab = "Bill Length (mm)",
     ylab="Body Mass (g)")

plot(df$species, df$body_mass_g, 
     xlab = "Species",
     ylab="Body Mass (g)")

plot(df$sex, df$body_mass_g, 
     xlab = "Sex",
     ylab="Body Mass (g)")



```

#(B)

#Performing a simple OLS linear regression of body mass (the response variable) on flipper length (the predictor)

```{r}
lm.fitted = lm(formula = body_mass_g ~ flipper_length_mm, data = df)
coefficients(lm.fitted)
```
The model is :
$$
\textrm{body mass} = -5872.0929 + 50.1533 * (\textrm{flipper length})
$$
#Is this a good model?
```{r}
Rsquared = summary(lm.fitted)$r.squared
```

```{r fig.width=5, fig.height=3}
qqnorm(lm.fitted$residuals)
```

##(C)

#Performing an OLS multiple linear regression of body mass on the four predictors: species, sex, flipper length and bill length.

```{r}
mlm.fitted = lm(formula = body_mass_g ~ ., data = df)
summary(mlm.fitted)
Rsquared = summary(mlm.fitted)$r.squared
```
#The model is:
$$
\textrm{body mass} = -759 
  - 291 * Chin 
  + 707 * Gen 
  + 465 * male
  + 17.8 * flipper
  + 21.6 * bill
$$

#Which of your variables have the strongest effect on body mass? Concretely, if you wanted to predict body mass, which other variable(s) would be most useful?

```{r}
dfz = df
dfz$flipper_length_mm = scale(dfz$flipper_length_mm)
dfz$bill_length_mm = scale(dfz$bill_length_mm)
dfz$body_mass_g = scale(dfz$body_mass_g) # optional
mlm.fitted = lm(formula = body_mass_g ~ ., data = dfz)
# summary(mlm.fitted) # to check p values
coefficients(mlm.fitted)
```
